A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-07-28 sections/03FeatureExtraction.Rmd
feedback lfb

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018, helmich2021}. Person level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in between-person differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or do they vary across the measurement period?''

@@

\paragraph{Instability.}
Instability captures the average change between two consecutive measurements \citep{ebner-priemer2009}. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the 'jumpy-ness' of the data over time. In other words, variability reflects the range or diversity of values in a time series data, while instability reflects the fluctuation or inconsistency in a time series data over time  \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Within psychological time series, instability measurements have especially been important in the research of borderline personality disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also in understanding early warning signals more generally \citep{wichers2019}. Conceptually, the instability feature, thus, relates to a broad range of research questions, including: "What is the nature of the identification changes in those who start working in a new country?" or "How does the pattern of day-to-day (i.e., measurement-to-measurement) fluctuations in self-esteem during adolescence reflect the process of identity formation and self-concept development?"

@@

```{=tex}
\begin{equation} \label{eq:mac}
  MAC(X_{ij}) = \frac{1}{n-1} \sum_{t=1, \ldots, t-1}\left|x_{t}-x_{t-1}\right|
\end{equation}
```

Another common measurement of instability is the \textit{Mean of the Squared Successive Differences} ($MSSD$), which is often preferred where differences in magnitude are more important than the frequency of those changes, for example, when big shifts in time series are considered more impactful or when outliers are meaningful and need to be taken into account.

@@

For our illustration case, we chose the commonly used autocorrelation or autoregression with a lag-1 to capture the inertia. High autocorrelation values can indicate high levels of inertia, while low autocorrelation values may indicate a more unpredictable or volatile time series \citep{dejonckheere2019}. The lag--1 autocorrelation $r_{ij,1}$ looks at the average correlation between a measurement $x$ and the proceeding measurement $x_{t-1}$ for the time series $X$ of participant $i$ and variable $j$.

```{=tex}
\begin{equation} \label{eq:ar}
  r_{ij,1} = \frac{\sum_{t=1}^{n-l}(x_{ijt}-\overline{x}_{ij})(x_{ij,t-1}-\overline{x}_{ij})}{\sum_{t=1}^{n}(x_{ijt}-\overline{x}_{ij})^2}
\end{equation}
```

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead, nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research questions about the type of the development: "Is the development of well-being a nonlinear process?" as well as the shape and structure of the development: "How many spikes in well-being did a migrant experience?" 
We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line by estimating the number of parametric and non-parametric (or smooth) terms $p$ in the model. An edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend.

```{=tex}
\begin{equation} \label{eq:edf}
  edf \approx \sum p
\end{equation}
```

>>>>

\paragraph{Variability.}
Variability captures the degree to which a set of data differs from the central tendency and is sometimes also referred to as the dispersion or spread of the data \citep{weisberg1992}. In time series analyses, variability is conceptually important because information about the distribution and diversity of the data has been found to be indicative of worse psychological states \citep{myin-germeys2018, helmich2021}. Person-level differences of ESM measurements have, for example, been associated with higher levels of psycho-pathological recurrences among depression patients \citep{timm2017}. As such, psychological researchers and practitioners are often empirically interested in between-person differences in variability. Researchers on polarization and radicalization might for example ask: "Are people settled in their attitudes towards migrants or do they vary across the measurement period?"

@@

\paragraph{Instability.}
Instability captures the average change between two consecutive measurements \citep{ebner-priemer2009}. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the 'jumpy-ness' of the data over time. In other words, variability reflects the range or diversity of values in a time series data, while instability reflects the fluctuation or inconsistency in a time series data over time  \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Within psychological time series, instability measurements have especially been important in the research of borderline personality disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also in understanding early warning signals more generally \citep{wichers2019}. Conceptually, the instability feature, thus, relates to a broad range of research questions, including: "What is the nature of the identification changes in those who start working in a new country?" or "Do strong daily fluctuations in self-esteem reflect the process of identity formation in adolescents?"

@@

```{=tex}
\begin{equation} \label{eq:mac}
  MAC(X_{ij}) = \frac{1}{n-1} \sum_{t=2, \ldots, t}\left|x_{t}-x_{t-1}\right|
\end{equation}
```

Another common measurement of instability is the \textit{Mean of the Squared Successive Differences} ($MSSD$), which is often preferred where differences in magnitude are more important than the frequency of those changes, for example, when big shifts in time series are considered more impactful or when outliers are meaningful and need to be taken into account \citep{chatfield2003}.

@@

For our illustration case, we chose the commonly used autocorrelation or autoregression with a lag-1 to capture the inertia. High autocorrelation values can indicate high levels of inertia, while low autocorrelation values may indicate a more unpredictable or volatile time series \citep{dejonckheere2019}. The lag--1 autocorrelation $r_{ij,1}$ looks at the average correlation between a measurement $x$ and the preceding measurement $x_{t-1}$ for the time series $X$ of participant $i$ and variable $j$ with $n$ measurements.

```{=tex}
\begin{equation} \label{eq:ar}
  r_{ij,1} = \frac{\sum_{t=2}^{n}(x_{ijt}-\overline{x}_{ij})(x_{ij,t-1}-\overline{x}_{ij})}{\sum_{t=1}^{n}(x_{ijt}-\overline{x}_{ij})^2}
\end{equation}
```

Where $\overline{x}_{ij}$ is the mean of the time series $x_{ij}$, calculated as:

```{=tex}
\begin{equation} \label{eq:mean_for_ar1}
  \overline{x}_{ij} = \frac{1}{n} \sum_{t=1}^{n} x_{ijt}
\end{equation}
```

@@

\paragraph{Nonlinearity.}
Changes in psychology are not always linear, instead, nonlinearity is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery from depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. Conceptually, researchers might have research questions about the type of the development: "Is the development of well-being a nonlinear process?" as well as the shape and structure of the development: "How many spikes in well-being did a migrant experience?"

We summarized the nonlinear trend with the \textit{estimated degrees of freedom} of an empty GAM spline model. The $edf$ summarizes the \textit{wiggliness} of a spline trend line \citep{wood2017}. The degrees of freedom of a spline model are primarily determined by the number of knots and the order of the spline. For instance, a cubic spline with $k$ knots has $k$+3 degrees of freedom \citep{faraway2016}. However, in a penalized spline framework, which is commonly used for GAMs, the effective degrees of freedom can be less than $k$+3. This is because the model employs a smoothing parameter to control the trade-off between the complexity (flexibility) of the model and its fit to the data, thereby penalizing overly complex models and potentially reducing the effective degrees of freedom \citep{marx1998}. Intuitively then an edf of 1 would be equivalent to a linear relationship (i.e., one linear slope parameter), whereas a higher edf (particularly an edf > 2) is indicative of a non-linear trend. The estimated degrees of freedom are commonly based on a concept called 'effective degrees of freedom' and can be represented as the trace $tr$ (i.e., the sum of the diagonal elements) of the smoother matrix $S$, a symmetric matrix that maps from the raw data to the smooth estimates \citep{wood2017}.

```{=tex}
\begin{equation} \label{eq:df}
  edf = tr(S)
\end{equation}
```

<!--Effective degrees of freedom for individual smooth terms are obtained by summing the Fii values corresponding to the coefficients βi of the smooth term (e.g., if the coefficients of a particular smooth are β5, . . . , β11 then the smooth’s EDF is $\sum_{i=5}^{11} = F_{ii}).--!>

<!-- An alternative approach to understanding the edf is to approximate the number of parametric and non-parametric (or smooth) terms $p$ in the penalized GAM. Both formulas are measuring the same concept (the model's complexity or flexibility) and both account for the penalization introduced to control overfitting in GAMs \citep{hastie1986}.

```{=tex}
\begin{equation} \label{eq:edf}
  edf \approx \sum p
\end{equation}
```
--!>

### 2023-07-28 sections/04FeatureReduction.Rmd
feedback lfb

To use the principal component analysis with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration, we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` principal component scores for each participant (i.e., the participants' coordinates in the reduced dimensional space; PC-scores).

>>>>

To use the PCA with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA uses linear transformations in such a way that the first component captures the most possible variance of the original data \citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}. The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated \citep{shlens2014}. In practice, this meant that the PCA decomposed the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90% of the original variance \citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}. For our illustration, we select the first `r pca_cutoff` principal components that explain 80% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`%). For the extracted principal components we save the `r pca_cutoff` principal component scores for each participant (i.e., the participants' coordinates in the reduced dimensional space; PC-scores).

### 2023-07-28 sections/07ClusterEvaluationInterpretation.Rmd
feedback lfb

In short, we find that the feature-based clustering discerned two meaningfully different groups of participants. We find an adaptive group (cluster 2) that reports higher well-being and more positive outgroup attitudes (\textit{median}) that are also stable over time (\textit{MAD}, \textit{MAC}) and tend to increase over the 30 day test period (\textit{linear trend}). This group also reported consistently more meaningful, need-fulfilling, and cooperative outgroup interactions (\textit{median}). This group with overwhelmingly positive experiences stands in contrast with a more detrimental group (cluster 1). This cluster, on average, reported much less positive, less meaningful, and less fulfilling interactions and interaction patterns (\textit{median}). This group also reported less positive outgroup attitudes, lower well-being, and more discrimination experiences (\textit{median}). At the same time, for members of this detrimental cluster conditions seemed to deteriorate over time (\textit{linear trend}), and there was generally less consistency in the experiences they were able to have (\textit{MAC}, \textit{MAD}, \textit{edf}).

To identify these patterns, we first inspect the clusters based on the average values of meaningful features (see \fgrref[A]{fig:clusterFeatVar}; \citealp{Kennedy2021}). We see that for some variables the features are generally stronger in separating the clusters. We, for example, see that the item on '\textit{how cooperative the interaction was}' distinguishes the two clusters across almost all seven features (except for the \textit{auto-correlation}, see \fgrref[A]{fig:clusterFeatVar}). Compare this to the '\textit{outgroup attitudes}' item where the differences between the clusters are much smaller for almost all features. We then inspect the clusters with a focus on the features (see \fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the variable focus, we can see more clearly that some features are better at distinguishing the clusters across variables. For example, \textit{MAD} and \textit{median} distinguish the two clusters across almost all variables (except for the item of whether the interaction was representative of the outgroup). These two features stand in stark contrast to other features, such as the \textit{lag-1 auto correlations}, which showed much smaller differences between the two clusters (see \fgrref[B]{fig:clusterFeatVar}). This offers some information on which features were most important in understanding the two extracted groups. Taking these two perspectives together, we can also focus on individual features or variables in particular. We, for example, see a strong difference in the average well-being, where participants in cluster 2 showed a much lower median well-being over the time series. At the same time, in terms of stability, both groups have virtually identical averages \textit{MAC} statistics for well-being (see \fgrref[A]{fig:clusterFeatVar}). There are, thus, variables and features that distinguish the clusters better than others and a combination of variables and features lets us explore meaningful group differences in more detail. In our case, we see that the central tendency, variability, and linear trend are best at distinguishing a group with mainly positive experiences (cluster 1) from a group with a more negative experience (cluster 2). We also see that our clusters line up with the past literature on the importance of focusing on simpler and more meaningful statistics \citep{bringmann2018c, eronen2021a}.

@@

Finally, we can also assess the clusters across other individual difference variables \citep[e.g.,][]{monden2022}. This out-of-feature comparison allows us to check for data artifacts, as well as check whether the developmental clusters are associated with important social markers and individual differences. To illustrate artifact checks, we added the number of ESM measurements into the comparison and find that the participants in the deterioration cluster (cluster 2) on average completed slightly more ESM surveys in general and reported on more intergroup interactions in particular (see $n$ in \fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we ensured that the general time frame and completion rates are similar for all participants and indeed the numbers in ESM measurements generally are largely similar (e.g., see $n$ for well-being and outgroup attitudes). However, the difference in the reported number of interactions might indicate either a clustering artifact or a meaningful difference. The higher average number of interactions in cluster 2 could, for example, indicate a clustering artifact if variances are more inflated due to the larger samples \citep{kogan2006}. At the same time, however, the difference in the number of experienced interactions might also indicate a meaningful difference, where the deteriorating cluster (cluster 2) on average reported more outgroup interactions, but these interactions were less voluntary, less meaningful, and less positive. Such a finding is in line with past research highlighting the role of negative intergroup interactions in explaining intergroup relations \citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}. To further illustrate the utility of assessing out-of-feature individual differences, we also compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). When looking at the group comparison, we find that participants in the deteriorating cluster (cluster 2) reported substantially higher levels of everyday discrimination (\fgrref[B]{fig:clusterFeatVar}). Thus, both intensive longitudinal (e.g., the sum of specific ESM measurements) and cross-sectional variables (e.g., general discrimination differences) that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.

@@

```{=tex}
\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. Subplot (B) shows the GAM spline for each cluster across the measurement occasions.}
\end{figure}
```

>>>>

In short, we find that the feature-based clustering discerned two meaningfully different groups of participants. We find an adaptive group (cluster 1) that reports higher well-being and more positive outgroup attitudes (\textit{median}) that are also stable over time (\textit{MAD}, \textit{MAC}) and tend to increase over the 30 day test period (\textit{linear trend}). This group also reported consistently more meaningful, need-fulfilling, and cooperative outgroup interactions (\textit{median}). This group with overwhelmingly positive experiences stands in contrast with a more detrimental group (cluster 2). This cluster, on average, reported much less positive, less meaningful, and less fulfilling interactions and interaction patterns (\textit{median}). This group also reported less positive outgroup attitudes, lower well-being, and more discrimination experiences (\textit{median}). At the same time, for members of this detrimental cluster (cluster 2) conditions seemed to deteriorate over time (\textit{linear trend}), and there was generally less consistency in the experiences they were able to have (\textit{MAC}, \textit{MAD}, \textit{edf}).

To identify these patterns, we first inspect the clusters based on the average values of meaningful features (see \fgrref[A]{fig:clusterFeatVar}; \citealp{Kennedy2021}). We see that for some variables the features are generally stronger in separating the clusters. We, for example, see that the item on '\textit{how cooperative the interaction was}' distinguishes the two clusters across almost all seven features (except for the \textit{auto-correlation}, see \fgrref[A]{fig:clusterFeatVar}). Compare this to the '\textit{outgroup attitudes}' item where the differences between the clusters are much smaller for almost all features. We then inspect the clusters with a focus on the features (see \fgrref[B]{fig:clusterFeatVar}). While this is the same data as for the variable focus, we can see more clearly that some features are better at distinguishing the clusters across variables. For example, \textit{MAD} and \textit{median} distinguish the two clusters across almost all variables (except for the item of whether the interaction was representative of the outgroup). These two features stand in stark contrast to other features, such as the \textit{lag-1 auto correlations}, which showed much smaller differences between the two clusters (see \fgrref[B]{fig:clusterFeatVar}). This offers some information on which features were most important in understanding the two extracted groups. Taking these two perspectives together, we can also focus on individual features or variables in particular. We, for example, see a strong difference in the average well-being, where participants in cluster 2 showed a much lower median well-being over the time series. At the same time, in terms of stability, both groups have virtually identical average \textit{MAC} statistics for well-being (see \fgrref[A]{fig:clusterFeatVar}). There are, thus, variables and features that distinguish the clusters better than others and a combination of variables and features lets us explore meaningful group differences in more detail. In our case, we see that the central tendency, variability, and linear trend are best at distinguishing a group with mainly positive experiences (cluster 1) from a group with a more negative experience (cluster 2). We also see that our clusters line up with the past literature on the importance of focusing on simpler and more meaningful statistics \citep{bringmann2018c, eronen2021a}.

@@

Finally, we can also assess the clusters across other individual difference variables \citep[e.g.,][]{monden2022}. This out-of-feature comparison allows us to check for data artifacts, as well as check whether the developmental clusters are associated with important social markers and individual differences. To illustrate artifact checks, we added the number of ESM measurements into the comparison and find that the participants in the deterioration cluster (cluster 2) on average completed slightly more ESM surveys in general and reported on more intergroup interactions in particular (see $n$ in \fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we ensured that the general time frame and completion rates are similar for all participants and indeed the numbers in ESM measurements generally are largely similar (e.g., see $n$ for well-being and outgroup attitudes). However, the difference in the reported number of interactions might indicate either a clustering artifact or a meaningful difference. The higher average number of interactions in cluster 2 could, for example, indicate a clustering artifact if variances are more inflated due to the larger samples \citep{kogan2006}. At the same time, however, the difference in the number of experienced interactions might also indicate a meaningful difference, where the deteriorating cluster (cluster 2) on average reported more outgroup interactions, but these interactions were less voluntary, less meaningful, and less positive. Such a finding is in line with past research highlighting the role of negative intergroup interactions in explaining intergroup relations \citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}.

To further illustrate the utility of assessing out-of-feature individual differences, we also compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). When looking at the group comparison, we find that participants in the deteriorating cluster (cluster 2) reported substantially higher levels of everyday discrimination (\fgrref[B]{fig:clusterFeatVar}). Thus, both intensive longitudinal (e.g., the sum of specific ESM measurements) and cross-sectional variables (e.g., general discrimination differences) that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.

@@

```{=tex}
\begin{figure}[!ht] %hbtp
  \caption{Cluster Group Comparisons over time}
  \label{fig:clusterTs}
  \centering\includegraphics[width=\textwidth]{figures/clusterTsComb.pdf}
  \caption*{Note: \\
  Subplot (A) displays the variable cluster means at every measurement occasion. Subplot (B) shows the GAM spline for each cluster across the measurement occasions. The thinner lines present all individual time series.}
\end{figure}
```
